PANCHANGE_analysis

### 2020-09-28 PANCHANGE_analysis.Rmd
added half-violin function

@@

@@

```{r gad gather to long format Time_points gad.sum_score}
dat.long.gad.sum <- dat %>%
  gather(
    key = "Time_point_raw",
    value = "gad.sum_score",
    all_of(gad.sum_score_cols)) %>%
  select(
    ID,
    Sample,
    Gender_collapsed,
    Ethnicity_collapsed,
    Disorder_hierarchical,
    Time_point_raw,
    gad.sum_score
  ) %>%
  mutate(Time_point = str_replace(Time_point_raw, "gad.sum_score_", "")) %>%
  select(-Time_point_raw)

@@

@@

```{r subsample for prepan-baseline analysis}
dat.prepan <- dat %>%
  filter(
    ((data_group_gad == "Baseline and retrospective data" |
      data_group_gad == "Prepandemic and baseline data" |
      data_group_gad == "All prepandemic, baseline and retrospective data") |
      (data_group_phq == "Baseline and retrospective data" |
      data_group_phq == "Prepandemic and baseline data" |
      data_group_phq == "All prepandemic, baseline and retrospective data") |
      (data_group_pcl == "Baseline and retrospective data" |
      data_group_pcl == "Prepandemic and baseline data" |
      data_group_pcl == "All prepandemic, baseline and retrospective data") |
      (data_group_ocir == "Baseline and retrospective data" |
      data_group_ocir == "Prepandemic and baseline data" |
      data_group_ocir == "All prepandemic, baseline and retrospective data")) &
      prepandemic_march_1_numeric == 1
  )

>>>>

# R functions
## Half-violin plot
```{r Half-violin plot}
source("https://raw.githubusercontent.com/datavizpyr/data/master/half_flat_violinplot.R")
```

@@

# Function for split violin plot
```{r Function for split violin plot}
GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin, 
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
  data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
  grp <- data[1, "group"]
  newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
  newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
  newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

  if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
    stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
      1))
    quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
    aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
    aesthetics$alpha <- rep(1, nrow(quantiles))
    both <- cbind(quantiles, aesthetics)
    quantile_grob <- GeomPath$draw_panel(both, ...)
    ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
  }
  else {
    ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
  }
})

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
```

@@

```{r gad gather to long format Time_points gad.sum_score}
dat.long.gad.sum <- dat %>%
  gather(
    key = "Time_point_raw",
    value = "gad.sum_score",
    all_of(gad.sum_score_cols)) %>%
  select(
    ID,
    Sample,
    Gender_collapsed,
    Ethnicity_collapsed,
    Disorder_hierarchical,
    Time_point_raw,
    gad.sum_score
  ) %>%
  mutate(Time_point = str_replace(Time_point_raw, "gad.sum_score_", "")) %>%
  select(-Time_point_raw) %>%
  add_column(Full_cohort = as.factor("Full cohort")) %>%
  mutate(Time_point_dup = Time_point)

@@

```{r}
summary_data <- dat.long.gad.sum %>%
  group_by(Time_point) %>%
  summarise(
    mean = mean(gad.sum_score, na.rm = T),
    min = mean(gad.sum_score, na.rm = T) - qnorm(0.975)*sd(gad.sum_score, na.rm = T)/sqrt(n()),
    max = mean(gad.sum_score, na.rm = T) + qnorm(0.975)*sd(gad.sum_score, na.rm = T)/sqrt(n()),
    sd_low = mean(gad.sum_score, na.rm = T) - sd(gad.sum_score, na.rm = T),
    sd_up = mean(gad.sum_score, na.rm = T) + sd(gad.sum_score, na.rm = T)
  )

summary_data
```

```{r}
dat.long.gad.sum %>%
  filter(Time_point != "base") %>%
  ggplot(mapping = aes(
    x = Full_cohort,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 15, linetype = "dashed", color = "gray") +
  geom_split_violin(
    trim = FALSE,
    alpha = 0.5
    ) +
 stat_summary(fun.y = mean,
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "pointrange",
              position = position_dodge(width = .25)) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )
```

```{r gad violin sum}
violin.gad.sum.prepan.retro <- dat.long.gad.sum %>%
  filter(Time_point != "base") %>%
  ggplot(mapping = aes(
    x = Full_cohort,
    y = gad.sum_score,
    fill = Time_point)) +
    geom_hline(yintercept = 5, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 15, linetype = "dashed", color = "gray") +
  geom_split_violin(
    trim = FALSE,
    alpha = 0.5
    ) +
 stat_summary(fun.y = mean,
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "pointrange",
              position = position_dodge(width = .25)) +
    labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )
violin.gad.sum.prepan.retro

violin.gad.sum.prepan.base <- dat.long.gad.sum %>%
  filter(Time_point != "retro") %>%
  ggplot(mapping = aes(
    x = Full_cohort,
    y = gad.sum_score,
    fill = Time_point)) +
    geom_hline(yintercept = 5, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 15, linetype = "dashed", color = "gray") +
  geom_split_violin(
    trim = FALSE,
    alpha = 0.5
    ) +
 stat_summary(fun.y = mean,
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "pointrange",
              position = position_dodge(width = .25)) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )
violin.gad.sum.prepan.base

violin.gad.sum.retro.base <- dat.long.gad.sum %>%
  filter(Time_point != "prepan") %>%
  ggplot(mapping = aes(
    x = Full_cohort,
    y = gad.sum_score,
    fill = Time_point)) +
    geom_hline(yintercept = 5, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 15, linetype = "dashed", color = "gray") +
  geom_split_violin(
    trim = FALSE,
    alpha = 0.5
    ) +
 stat_summary(fun.y = mean,
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "pointrange",
              position = position_dodge(width = .25)) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )
violin.gad.sum.retro.base

violin.gad.sum.prepan.retro +
  violin.gad.sum.prepan.base +
  violin.gad.sum.retro.base +
  plot_layout(guides = "collect", widths = c(1,1,1)) +
  plot_annotation(tag_levels = "A")
```

```{r gad violin sum}
dat.long.gad.sum %>%
  filter(Time_point != "base") %>%
  ggplot(mapping = aes(
    x = Sample,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_split_violin(
#    color = textcolor,
    trim = FALSE,
    alpha = 0.5
    ) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )+
  coord_flip()
```

```{r gad violin sum}
dat.long.gad.sum %>%
  filter(Time_point != "base") %>%
  ggplot(mapping = aes(
    x = Disorder_hierarchical,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_split_violin(
#    color = textcolor,
    trim = FALSE,
    alpha = 0.5
    ) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )+
  coord_flip()
```

```{r gad violin sum}
dat.long.gad.sum %>%
  filter(Time_point != "retro") %>%
  ggplot(mapping = aes(
    x = Disorder_hierarchical,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_split_violin(
#    color = textcolor,
    trim = FALSE,
    alpha = 0.5
    ) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )+
  coord_flip()
```

```{r gad violin sum}
dat.long.gad.sum %>%
  filter(Time_point != "prepan") %>%
  ggplot(mapping = aes(
    x = Disorder_hierarchical,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_split_violin(
#    color = textcolor,
    trim = FALSE,
    alpha = 0.5
    ) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    )+
  coord_flip()
```

```{r gad violin sum}
dat.long.gad.sum %>%
  filter(Sample == "GLAD") %>%
  filter(Time_point != "base") %>%
  ggplot(mapping = aes(
    x = Disorder_hierarchical,
    y = gad.sum_score,
    fill = Time_point)) +
  geom_split_violin(
#    color = textcolor,
    trim = FALSE,
    alpha = 0.5
    ) +
  labs(
    x = "GAD7 scores",
    y = "Percentage of participants",
    fill = "Time point") +
  theme_personal +
  scale_fill_manual(
    values = COPINGpalette3,
    breaks = c("prepan", "retro", "base"),
    labels = c("base" = "Baseline", "prepan" = "Prepandemic", "retro" = "Retrospective")
    ) +
  coord_flip()
```

```{r}
data %>%
  ggplot(aes(sex, dv, fill = face_sex)) +
  geom_hline(yintercept=0, color=textcolor, size=1) +
  geom_split_violin(color=textcolor, trim=FALSE, alpha = 0.5) +
  geom_boxplot(color = textcolor, 
               width = 0.25, 
               position = position_dodge(width=0.25)) +
  ylab("DV") +
  xlab("Participant Sex") +
  ylim(-5, 5) + 
  scale_fill_manual(values = c("#3D66CC", "#892110")) +
  my_theme
```

@@

```{r subsample for prepan-baseline analysis}
dat.prepan <- dat %>%
  filter(
    ((data_group_gad == "Baseline and retrospective data" |
      data_group_gad == "Prepandemic and baseline data" |
      data_group_gad == "All prepandemic, baseline and retrospective data") |
      (data_group_phq == "Baseline and retrospective data" |
      data_group_phq == "Prepandemic and baseline data" |
      data_group_phq == "All prepandemic, baseline and retrospective data") |
      (data_group_pcl == "Baseline and retrospective data" |
      data_group_pcl == "Prepandemic and baseline data" |
      data_group_pcl == "All prepandemic, baseline and retrospective data") |
      (data_group_ocir == "Baseline and retrospective data" |
      data_group_ocir == "Prepandemic and baseline data" |
      data_group_ocir == "All prepandemic, baseline and retrospective data")) &
      prepandemic_march_1_numeric == 1 # Exclude participants that signed up after the 1. March
  )
